\thetitle

### 2020-11-13 sun_phases/spin_qudit_tomo.tex
minor changes to appendix on reconstruction error, and changes to the corresponding section in the main text

\newcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\I}{\mathcal{I}}
\renewcommand{\L}{\mathcal{L}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\Q}{\mathcal{Q}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\T}{\mathcal{T}}

@@

The assumption that observables can be estimated up to uncorrelated noise with maximal variance $\epsilon^2$, summarized by Eq.~\eqref{eq:error_assumption}, is reasonable when measurement error is dominated by experimental sources of noise.
However, this assumption breaks down when measurement error is limited by fundamental quantum (shot) noise.
We relax the assumption of Eq.~\eqref{eq:error_assumption} and derive more fundamental error bounds in Appendix \ref{sec:error}.
Therein, we find that if measurement error is dominated by shot noise, then the estimate $\tilde\rho_V$ built from $n$ independent measurements of spin projection along every axis $\v v\in V$ approximates the true state $\rho$ to within a mean squared distance $\epsilon_V^2/n$, i.e.
\begin{align}
  \Bbk{\norm{\tilde\rho_V-\rho}^2} < \f{\epsilon_V^2}{n},
  \label{eq:bound}
\end{align}
where the {\it error scale} $\epsilon_V^2$ is
\begin{align}
  \epsilon_V^2 \equiv
  \f1d \sum_{\ell>0} \p{2\ell+1} \gamma_\ell^2 \S_{V\ell}^2,
  \label{eq:scale}
\end{align}
and $\gamma_\ell$ is defined using Clebsch-Gordan coefficients:
\begin{align}
  \gamma_\ell
  \equiv \f{\max_\mu\tau_{\ell\mu} - \min_\mu\tau_{\ell\mu}}{2},
  &&
  \tau_{\ell\mu} \equiv \bk{s\mu;\ell,0|s\mu}.
\end{align}
If $d$ even or $\ell$ is odd, then $\gamma_\ell=\max_\mu\tau_{\ell\mu}$.
For comparison with the previous bound in Eq.~\eqref{eq:bound_eps}, the error scale $\epsilon_V^2<\S_V^2/2$, so the previous bound still holds with the replacement $\epsilon^2\to1/2n$.
We note that the bound in Eq.~\eqref{eq:bound} is not tight, as it is acquired by bounding the statistical variance $\epsilon_{\v v\ell}^2$ in the empirical estimate of $\bk{T_{\v v\ell,0}}_\rho$ as $\bbk{\epsilon_{\v v\ell}^2}\le\frac{2\ell+1}{nd}\times\gamma_\ell^2$.
Though the individual bounds for each axis $\v v$ and degree $\ell$
are tight, these bounds cannot all be achieved simultaneously.
There is therefore still room for improvement on the bound in Eq.~\eqref{eq:bound}.

As a final comment, we note that Eq.~\eqref{eq:bound} provides a pessimistic upper bound on statistical error, which can be calculated without prior knowledge of the true qudit state $\rho$.
The actual error in the reconstruction $\tilde\rho_V$ of a particular state $\rho$ may be considerably smaller.
To write out this error in full (see Appendix \ref{sec:error} for a derivation), we define the $\p{2\ell+1}\times\p{2\ell+1}$ matrices
\begin{align}
  \N_{V\ell} &\equiv M_{V\ell}
  \diag\sp{M_{V\ell}^{-1} \p{M_{V\ell}^{-1}}^\dag} M_{V\ell}^\dag, \\
  \T_{\rho\ell} &\equiv \sum_{m,n}
  \cov_\rho\p{T_{\ell m}^\dag, T_{\ell n}} \op{m}{n},
\end{align}
where $\diag\sp{X}$ sets all off-diagonal entries of $X$ to zero, and $\cov_\rho\p{X,Y} \equiv \bk{XY}_\rho - \bk{X}_\rho \bk{Y}_\rho$.
We discuss details of efficiently computing $\N_{V\ell}$ in Appendix \ref{sec:compute}.
These matrices allow us to compactly express
\begin{align}
  \Bbk{\norm{\tilde\rho_V-\rho}^2}
  = \f1n \sum_\ell \p{\N_{V\ell}|\T_{\rho\ell}}.
  \label{eq:error}
\end{align}
While the error in Eq.~\eqref{eq:error} cannot be known exactly without knowing $\rho$, this error can be estimated by computing the right-hand side of Eq.~\eqref{eq:error} with the replacement $\rho\to\tilde\rho_V$.

[\red{I think the error in Eq.~\eqref{eq:error} can be maximized with no added cost to computational complexity.
  If so, the second half of this section entire section (as well as Appendix \ref{sec:error}) should be rewritten to replace Eq.~\eqref{eq:bound} by the tighter bound obtained by maximizing Eq.~\eqref{eq:error}.}]

@@

  \T_{\rho\ell} \equiv

@@

  = \f1n \sum_\ell \obk{\N_{V\ell}|\T_{\rho\ell}}.

@@

@@

We now seek to maximize the error in Eq.~\eqref{eq:error_SM} over all $\rho$, which will provide an upper bound on reconstruction error.
To this end, we vary the error with respect to $\rho$ by considering the vector derivative (i.e.~gradient)
  \f{\partial}{\partial\obra\rho} \Bbk{\norm{\tilde\rho_V-\rho}^2}
  = \f1n \sum_\ell \tr\p{\N_{V\ell}
    \f{\partial}{\partial\obra\rho} \T_{\rho\ell}}
  = 0,
  \f{\partial}{\partial\obra\rho}
  \equiv \sum_{L,M} \oket{T_{LM}} \f{\partial}{\partial\rho_{LM}}.
  \label{eq:variation}
Here $\partial/\partial\obra\rho$ is essentially a vector of all $\partial/\partial\rho_{LM}$, which satisfies
  \f{\partial}{\partial\obra\rho} \obk{\rho|X}
  = \f{\partial}{\partial\obra\rho}
  \sum_{\ell,m} \obk{\rho|T_{\ell m}} \obk{T_{\ell m}|X}
  = \sum_{L,M} \oket{T_{LM}} \f{\partial}{\partial\rho_{LM}}
  \sum_{\ell,m} \rho_{\ell m} \obk{T_{\ell m}|X}
  = \sum_{L,M} \oket{T_{LM}} \obk{T_{LM}|X}
  = \oket{X}.
The variation in Eq.~\eqref{eq:variation} leads us to consider
  \f{\partial}{\partial\obra\rho} \T_{\rho\ell}
  &= \sum_{m,n} \op{m}{n} \otimes \f{\partial}{\partial\obra\rho}
  \sp{\bk{T_{\ell m}^\dag T_{\ell n}}_\rho
    - \bk{T_{\ell m}^\dag}_\rho \bk{T_{\ell n}}_\rho} \\
  &= \sum_{m,n} \op{m}{n} \otimes \f{\partial}{\partial\obra\rho}
  \sp{\obk{\rho|T_{\ell m}^\dag T_{\ell n}}
    - \obk{\rho|T_{\ell m}^\dag} \obk{\rho|T_{\ell n}}} \\
  &= \sum_{m,n} \op{m}{n} \otimes \sp{\oket{T_{\ell m}^\dag T_{\ell n}}
    - \oket{T_{\ell m}^\dag} \obk{\rho|T_{\ell n}}
    - \obk{\rho|T_{\ell m}^\dag} \oket{T_{\ell n}}}.
  \label{eq:variation_cov}
The $\sim\rho$ terms in Eq.~\eqref{eq:variation_cov} can be simplified by the fact that
  \obk{\rho|T_{\ell m}}
  = \tr\p{\rho T_{\ell m}}
  = \p{-1}^m \tr\p{T_{\ell,-m}^\dag \rho}
  = \p{-1}^m \obk{T_{\ell,-m}|\rho}
  = \obk{T_{\ell m}^\dag|\rho},
  \sum_{m,n} \op{m}{n} \otimes
  \sp{\oket{T_{\ell m}^\dag} \obk{\rho|T_{\ell n}}
    + \obk{\rho|T_{\ell m}^\dag} \oket{T_{\ell n}}}
  &= \sum_{m,n} \op{m}{n} \otimes
  \sp{\oket{T_{\ell m}^\dag} \obk{T_{\ell n}^\dag|\rho}
    + \oket{T_{\ell n}} \obk{T_{\ell m}|\rho}} \\
  &= \sp{\sum_{m,n} \p{\I\op{m}{n}\I + \op{n}{m}}
    \otimes \oop{T_{\ell m}}{T_{\ell n}}} \oket{\rho},
where we define the inversion operator $\I\equiv\sum_m \p{-1}^m\op{-m}{m}$.
Meanwhile, the constant terms in Eq.~\eqref{eq:variation_cov} are
  \sum_{m,n} \op{m}{n} \otimes \oket{T_{\ell m}^\dag T_{\ell n}}
  = \sum_{L,m,n} \I \op{m}{n} \otimes g_{\ell mn}^L \oket{T_{L,m+n}},
  g_{\ell mn}^L \equiv \obk{T_{L,m+n}|T_{\ell m} T_{\ell n}}
  = f_{\ell m;\ell n}^{L,m+n}
Altogether, we have that
  \f{\partial}{\partial\obra\rho} \T_{\rho\ell}
  = \sum_{m,n} g_{\ell mn}^L \I\op{m}{n} \otimes \oket{T_{L,m+n}}
  - \sp{\sum_{m,n} \p{\op{n}{m} + \I\op{m}{n}\I}
    \otimes \oop{T_{\ell m}}{T_{\ell n}}} \oket{\rho}.
Substituting this result back into Eq.~\eqref{eq:variation}, we find that
  n\, \f{\partial}{\partial\obra\rho}
  = \oket{\chi_V} - \sum_\ell \L_{V\ell} \oket\rho
  = 0,
  \oket{\chi_V} \equiv \sum_{L,M} \chi^V_{LM} \oket{T_{LM}},
  \L_{V\ell} \equiv \sum_{m,n}
  \sp{\bk{m|\N_{V\ell}|n} + \bk{n|\I\N_{V\ell}\I|m}}
  \oop{T_{\ell m}}{T_{\ell n}}.
  \chi^V_{LM} \equiv \sum_\ell \obk{\N_{V\ell}|\D_M|\I g_{L\ell}},
  g_{L\ell} \equiv \sum_{m,n} g^L_{\ell mn} \op{m}{n},
  \D_M \equiv \sum_{m,n} \delta_{M,m+n} \op{mn}{mn}.
Defining the projection of $\oket{\chi_V}$ onto the degree-$\ell$ subspace of transition operators,
  \oket{\chi_{V\ell}}
  \equiv \sum_m \oket{T_{\ell m}} \obk{T_{\ell m}|\chi_V},
we thus find that reconstruction error is maximized by the state
  \oket\rho = \sum_\ell \L_{V\ell}^{-1} \oket{\chi_{V\ell}},
where $\L_{V\ell}^{-1}$ is the inverse of $\L_{V\ell}$.

[\red{todo: substitute this solution back into the reconstruction error to find an upper bound}]

>>>>

\newcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\I}{\mathcal{I}}
\renewcommand{\L}{\mathcal{L}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\Q}{\mathcal{Q}}
\renewcommand{\S}{\mathcal{S}}

@@

The assumption that observables can be estimated up to uncorrelated noise with maximal variance $\epsilon^2$, summarized by Eq.~\eqref{eq:error_assumption}, is reasonable when measurement error is dominated by experimental sources of noise.
However, this assumption breaks down when measurement error is limited by fundamental quantum (shot) noise.
We relax the assumption of Eq.~\eqref{eq:error_assumption} in Appendix \ref{sec:error}, where we instead assume that $\tilde\rho_V$ is built from $n$ independent measurements of spin projection along every axis $\v v\in V$.
In this case, the mean squared distance between the approximation $\tilde\rho_V$ and the true state $\rho$ can be compactly expressed as as an inner product of the {\it covariance matrices}
\begin{align}
  \C_\ell\sp{\rho} \equiv \sum_{m,n}
  \cov_\rho\p{T_{\ell m}^\dag, T_{\ell n}} \op{m}{n}
\end{align}
where $\cov_\rho\p{X,Y} \equiv \bk{XY}_\rho - \bk{X}_\rho \bk{Y}_\rho$, and the {\it shot noise matrices}
\begin{align}
  \N_{V\ell} \equiv M_{V\ell} \diag\sp{M_{V\ell}^{-1}
    \p{M_{V\ell}^{-1}}^\dag} M_{V\ell}^\dag,
\end{align}
where $\diag\sp{X}$ sets all off-diagonal parts of $X$ to zero:
\begin{align}
  \Bbk{\norm{\tilde\rho_V-\rho}^2}
  = \f1n \sum_\ell \obk{\N_{V\ell}|\C_\ell\sp{\rho}}.
  \label{eq:error}
\end{align}
This reconstruction error can be maximized analytically, yielding a reconstruction error bound of the form
\begin{align}
  \Bbk{\norm{\tilde\rho_V-\rho}^2} \le \f{\epsilon_V^2}{n},
  \label{eq:bound}
\end{align}
where the complexity of computing the {\it error scale} $\epsilon_V^2$ is the same as that of computing $\S_V^2$.
For comparison with the bound in Eq.~\eqref{eq:bound_eps}, we note that the error scale $\epsilon_V^2<\S_V^2/2$, so the previous bound still holds with the replacement $\epsilon^2\to1/2n$.

[\red{todo: finish discussion about $\epsilon_V^2$}]

As a final comment, we note that Eq.~\eqref{eq:bound} only provides a pessimistic upper bound on statistical error, which can be calculated without prior knowledge of the true qudit state $\rho$.
The actual error in the reconstruction $\tilde\rho_V$ of a particular state $\rho$, provided by Eq.~\eqref{eq:error}, may be considerably smaller.
While the true error in $\tilde\rho_V$ cannot be known exactly without knowing $\rho$, this error can be estimated by computing the right-hand side of Eq.~\eqref{eq:error} with the replacement $\rho\to\tilde\rho_V$.

@@

  \C_\ell\sp{\rho} \equiv

@@

  = \f1n \sum_\ell \obk{\N_{V\ell}|\C_\ell\sp{\rho}}.

@@

In practice, to evaluate this error it is useful to expand the covariance matrix as
  \C_\ell\sp{\rho}
  = \sum_{m,n} \op{m}{n} \sp{\obk{\rho|T_{\ell m}^\dag T_{\ell n}}
    - \obk{\rho|T_{\ell m}^\dag} \obk{\rho|T_{\ell n}}}
  = \sum_{m,n} \I\op{m}{n} \sp{\obk{\rho|T_{\ell m} T_{\ell n}}
    - \obk{\rho|T_{\ell m}} \obk{\rho|T_{\ell n}}},
where we define the inversion operator $\I\equiv\sum_m \p{-1}^m\op{-m}{m}$, and then expand the product $T_{\ell m} T_{\ell n}$ to get
  \C_\ell\sp{\rho} = \sum_{m,n} \I \op{m}{n}
  \sp{\sum_L g_{\ell mn}^L \obk{\rho|T_{L,m+n}}
    - \obk{\rho|T_{\ell m}} \obk{\rho|T_{\ell n}}},
  g_{\ell mn}^L \equiv \obk{T_{L,m+n} | T_{\ell m} T_{\ell n}}
  = f_{\ell m;\ell n}^{L,m+n},

@@

We now seek to maximize the error in Eq.~\eqref{eq:error_SM} over all $\rho$, which will provide an upper bound on reconstruction error.
To this end, we vary the error with respect to $\rho$, and consider the vector derivative (i.e.~gradient)
\begin{align}
  \f{\partial}{\partial\obra\rho} \Bbk{\norm{\tilde\rho_V-\rho}^2}
  = \f1n \sum_\ell \tr\p{\N_{V\ell}
    \f{\partial}{\partial\obra\rho} \C_\ell\sp{\rho}},
  &&
  \f{\partial}{\partial\obra\rho}
  \equiv \sum_{L,M} \oket{T_{LM}} \f{\partial}{\partial\rho_{LM}}.
  \label{eq:variation}
\end{align}
Here $\partial/\partial\obra\rho$ is essentially a vector of all $\partial/\partial\rho_{LM}$, which satisfies
\begin{align}
  \f{\partial}{\partial\obra\rho} \obk{\rho|X}
  = \f{\partial}{\partial\obra\rho}
  \sum_{\ell,m} \obk{\rho|T_{\ell m}} \obk{T_{\ell m}|X}
  = \sum_{L,M} \oket{T_{LM}} \f{\partial}{\partial\rho_{LM}}
  \sum_{\ell,m} \rho_{\ell m} \obk{T_{\ell m}|X}
  = \sum_{L,M} \oket{T_{LM}} \obk{T_{LM}|X}
  = \oket{X}.
\end{align}
The variation in Eq.~\eqref{eq:variation} leads us to consider
\begin{align}
  \f{\partial}{\partial\obra\rho} \C_\ell\sp{\rho}
  &= \sum_{m,n} \I \op{m}{n}
  \f{\partial}{\partial\obra\rho}
  \sp{\sum_L g_{\ell mn}^L \obk{\rho|T_{L,m+n}}
    - \obk{\rho|T_{\ell m}} \obk{\rho|T_{\ell n}}} \\
  &= \sum_{m,n} \I\op{m}{n} \otimes
  \sp{\sum_L g_{\ell mn}^L \oket{T_{L,m+n}}
    - \oket{T_{\ell m}} \obk{\rho|T_{\ell n}}
    - \obk{\rho|T_{\ell m}} \oket{T_{\ell n}}}.
  \label{eq:variation_cov}
\end{align}
The $\sim\rho$ terms in Eq.~\eqref{eq:variation_cov} can be simplified by the fact that
\begin{align}
  \obk{\rho|T_{\ell m}}
  = \tr\p{\rho T_{\ell m}}
  = \p{-1}^m \tr\p{T_{\ell,-m}^\dag \rho}
  = \p{-1}^m \obk{T_{\ell,-m}|\rho}
  = \obk{T_{\ell m}^\dag|\rho},
\end{align}
which implies that
\begin{align}
  \sum_{m,n} \I \op{m}{n} \otimes
  \sp{\oket{T_{\ell m}} \obk{\rho|T_{\ell n}}
    + \obk{\rho|T_{\ell m}} \oket{T_{\ell n}}}
  &= \sum_{m,n} \I \op{m}{n} \otimes
  \sp{\oket{T_{\ell m}} \obk{T_{\ell n}^\dag|\rho}
    + \oket{T_{\ell n}} \obk{T_{\ell m}^\dag|\rho}} \\
  &= \sp{\sum_{m,n} \p{\I\op{m}{n}\I + \op{n}{m}}
    \otimes \oop{T_{\ell m}}{T_{\ell n}}} \oket{\rho}.
\end{align}
Altogether, we have that
\begin{align}
  \f{\partial}{\partial\obra\rho} \C_\ell\sp{\rho}
  = \sum_{m,n} g_{\ell mn}^L \I\op{m}{n} \otimes \oket{T_{L,m+n}}
  - \sp{\sum_{m,n} \p{\op{n}{m} + \I\op{m}{n}\I}
    \otimes \oop{T_{\ell m}}{T_{\ell n}}} \oket{\rho},
\end{align}
which we can substitute back into Eq.~\eqref{eq:variation} to find that
\begin{align}
  n\, \f{\partial}{\partial\obra\rho}
  \Bbk{\norm{\tilde\rho_V-\rho}^2}
  = \oket{\chi_V} - \sum_\ell \L_{V\ell} \oket\rho,
\end{align}
where
\begin{align}
  \oket{\chi_V} \equiv \sum_{L,M} \chi^V_{LM} \oket{T_{LM}},
  &&
  \L_{V\ell} \equiv \sum_{m,n}
  \sp{\bk{m|\N_{V\ell}|n} + \bk{n|\I\N_{V\ell}\I|m}}
  \oop{T_{\ell m}}{T_{\ell n}}.
\end{align}
\begin{align}
  \chi^V_{LM} \equiv \sum_\ell \obk{\N_{V\ell}|\D_M|\I g_{L\ell}},
  &&
  g_{L\ell} \equiv \sum_{m,n} g^L_{\ell mn} \op{m}{n},
  &&
  \D_M \equiv \sum_{m,n} \delta_{M,m+n} \op{mn}{mn}.
\end{align}
Defining the projection of $\oket{\chi_V}$ onto the degree-$\ell$ subspace of transition operators,
\begin{align}
  \oket{\chi_{V\ell}}
  \equiv \sum_m \oket{T_{\ell m}} \obk{T_{\ell m}|\chi_V},
\end{align}
we thus find that reconstruction error is maximized by the state
\begin{align}
  \oket{\rho^\star} = \sum_\ell \L_{V\ell}^{-1} \oket{\chi_{V\ell}},
\end{align}
where $\L_{V\ell}^{-1}$ is the inverse of $\L_{V\ell}$.
In total, the ultimate upper bound on reconstruction error is
\begin{align}
  \Bbk{\norm{\tilde\rho_V-\rho}^2} \le
  \left. \Bbk{\norm{\tilde\rho_V-\rho}^2} \right|_{\rho=\rho^\star}
  = \f1n \sum_{\ell} \obk{\N_{V\ell}|\C_\ell\sp{\rho^\star}}.
\end{align}

[\red{todo: finish substituting $\rho^\star$, or otherwise discuss an ``efficient'' calculation of this bound.}]
